Dynamics of Colombo's Top: Tidal Dissipation and Resonance Capture

### 2021-05-04 writeups/Su_weak_tides.tex
can't decide on color but technically have a draft

In this section, we consider a simplified model of equilibrium tides that
isolates the important new phenomenon presented in this paper. We assume that
the spin magnitude of the planet is constant, so $\alpha$ and $g$ are both
fixed, while the spin orientation $\uv{s}$ experiences an alignment torque
towards $\uv{l}$ on the alignment timescale $t_{\rm s}$:
\begin{equation}
    \p{\rd{\uv{s}}{t}}_{\rm tide}
        = \frac{1}{t_{\rm s}} \uv{s} \times \p{\uv{l} \times \uv{s}}.
        \label{eq:dsdt_tide_toy}
\end{equation}
The full equations of motion for $\uv{s}$ in the coordinates $\theta$ and $\phi$
can be written:
\begin{align}
    \rd{\theta}{t} &= -g\sin I \sin \phi - \frac{1}{t_{\rm s}} \sin \theta,
        \label{eq:dqdt_toy}\\
    \rd{\phi}{t} &= -\alpha \cos\theta
        - g\p{\cos I + \sin I \cot \theta \cos \phi}.\label{eq:dfdt_toy}
\end{align}

@@

If the alignment torque is weak ($\abs{gt_{\rm s}} \gg 1$), then the fixed points of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) are just slightly shifted CSs. This
shift can be calculated cleanly to leading order: all of the CS obliquities
$\theta_{\rm cs}$ are unchanged while the azimuthal angle $\phi_{\rm cs}$ for
each CS now satisfies
\begin{equation}
    \sin \phi_{\rm cs} = \frac{\sin\theta_{\rm cs}}{\sin I \abs{g}t_{\rm s}}.
        \label{eq:mcs_shift}
\end{equation}
We can see that if $t_{\rm s} > t_{\rm s, c}$, where
\begin{align}
    t_{\rm s, c} &= \frac{1}{\abs{g}\sin I},\label{eq:mcs_shift_crit}
\end{align}
then Eq.~\eqref{eq:mcs_shift} will always have solutions for $\phi_{\rm cs}$,
and the alignment torque never changes the number of fixed points of the system.
If $t_{\rm s}$ is decreased below $t_{\rm s, c}$, CS2 and CS4 cease to be
fixed points if $\eta \ll 1$ \citep[as first noted in][]{fabrycky_otides}, as
$\theta_{\rm cs} \approx 90^\circ$ for these (see Fig.~\ref{fig:cs_locs}), while
the other CSs have small $\sin \theta_{\rm cs}$ and are only slightly perturbed.
Figure~\ref{fig:mcs} shows the obliquity and azimuthal angles for each of the
CSs in the $\eta \ll 1$ case, where it can be seen that CS2 and CS4 collide and
annihilate. The phase shift $\phi_{\rm cs}$ for CS2 and CS4 for $t_{\rm s} >
t_{\rm s, c}$ can be predicted to good accuracy using Eq.~\eqref{eq:mcs_shift}
using $\theta_{\rm cs} \approx \pi/2 - \eta \cos (I) \approx 79^\circ$
\citep{su2020} for both, shown as the dashed lines in the bottom panel of
Fig.~\ref{fig:mcs}. For the remainder of this section, we will consider the case
where $t_{\rm s} \gg t_{\rm s, c}$ and the CSs only differ slightly from their
unperturbed locations.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/0_stab.png}
    \caption{Modified CS obliquities (top) and azimuthal angles (bottom) for $I
    = 20^\circ$ and $\eta = 0.2$, where the CS1 and CS3 obliquities have been
    offset (labeled in top legend) to improve clarity of the plot. In both
    panels, the solid lines give the result when applying a numerical root
    finding algorithm to the full equations of motion,
    Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}), while the dotted lines in the
    bottom panel give the CS2 and CS4 azimuthal angles according to
    Eq.~\eqref{eq:mcs_shift}. At $\abs{gt_{\rm s} \sin I} = 1$, CS2 and CS4
    collide and annihilate (Eq.~\ref{eq:mcs_shift_crit}).}\label{fig:mcs}
\end{figure}

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the weak tidal alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a shifted CS, yelding
\begin{align}
    \rd{}{t}\begin{bmatrix}
        \theta\\ \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm s}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and $\Delta \phi = \phi - \phi_{\rm cs}$. The eigenvalues
$\lambda$ of Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{equation}
    0 = \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm s}}}\lambda
        - \lambda_0^2,\label{eq:lambda_orig}
\end{equation}
where
\begin{equation}
    \lambda_0^2 \equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}}\label{eq:def_l0_sq}.
\end{equation}
When $t_{\rm s}$ is large, we can simplify Eq.~\eqref{eq:lambda_orig} to
\begin{equation}
    \lambda \approx -\frac{\cos \theta_{\rm cs}}{t_{\rm s}}
        \pm \sqrt{\lambda_0^2}. \label{eq:lambda_approx}
\end{equation}
The stability depends only on the real part of $\lambda$ in
Eq.~\eqref{eq:lambda_approx}. $\lambda_0^2$ is a generalization of Eq.~(A4)
in, \citet{su2020} and generally has the same behavior: it is negative for CSs
1--3 and positive for CS4, as shown in Fig.~\ref{fig:lambda_full}. Thus, CS4 is
always unstable, as there will always be at least one positive solution for
$\lambda$, and the stability of CSs 1--3 are solely determined by the sign of
$\cos \theta_{\rm cs}$. Evaluating at each of the CSs (see
Fig.~\ref{fig:cs_locs}), we conclude that CS1 and CS2 are stable and attracting
while CS3, stable in the absence of the torque, becomes unstable. These
calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal, fabrycky_otides}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_lambdas_full.png}
    \caption{$\lambda_0^2$ (Eq.~\ref{eq:def_l0_sq}) as a function of $\eta$ for
    the four CSs. The solid lines give $\lambda_0^2$ for the unperturbed
    $\phi_{\rm cs}$, and the dashed lines give the values for $\phi_{\rm cs}$
    shifted by $60^\circ$ ($\phi_{\rm cs} = 120^\circ$ for CS2 and $\phi_{\rm
    cs} = 60^\circ$ for CSs 1, 3, and 4).
    }\label{fig:lambda_full}
\end{figure}

@@

With the above results, we are equipped to ask questions about the dynamics of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}): what is the long term behavior for
a general initial $\uv{s}$? If $\eta > \eta_{\rm c}$, then the only possible
final spin state is CS2, and all initial conditions will evolve asympotically
towards it. As such, we consider only $\eta < \eta_{\rm c}$, where an initial
condition can asymptotically evolve towards either CS1 or CS2. We numerically
integrate Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) for many random initial
conditions uniformly distributed in $\p{\cos \theta, \phi}$ with $\abs{g}t_{\rm
s} = 10^{-3}$ and record the nearest CS for each integration after $10t_{\rm
s}$. In Fig.~\ref{fig:toy_phop}, we show the results of this procedure for $\eta
= 0.2$, and $I = 20^\circ$. It is clear that initial conditions in zone I evolve
into CS1, those in zone II evolve into CS2, while those in zone III have a
probabilistic outcome. We aim to understand each of these in turn:
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_2.png}
    \caption{Plot illustrating the asymptotic behavior of initial conditions for
    $\eta = 0.2$ and $I = 20^\circ$. Each dot represents an initial spin
    orientation, and the coloring of the dot indicates which Cassini State
    (legend) the system asymptotes towards.}\label{fig:toy_phop}
\end{figure}

@@

To understand this process quantitatively, as well as associate probabilities to
the two possible outcomes, it is important to be more quantitative. The correct
approach is to consider the evolution of the value of the \emph{unperturbed}
Hamiltonian (Eq.~\ref{eq:H}) as the spin evolves with the alignment torque. A
point in zone III evolves such that $H$ is increasing until $H \approx H_{\rm
sep}$ where $H_{\rm sep}$ is the value of $H$ along the separatrix, given by
\begin{align}
    H_{\rm sep} &\equiv H\p{\cos \theta_{\rm 4}, \phi_{\rm 4}},\nonumber\\
        &\approx g\sin I + \frac{g^2}{2\alpha}\cos^2 I +
            \mathcal{O}\p{\eta^2},\label{eq:def_Hsep}
\end{align}
where $\theta_4$ and $\phi_4$ are the coordinates of CS4. As the system evolves
closer to the separatrix, the change in $H$ over each circulation cycle can be
approximated by $\Delta H_-$, the change in $H$ along $\mathcal{C}_-$ (see
Fig.~\ref{fig:1contours}). In general, we can compute $\Delta H_{\pm}$ the
change over both legs of the separatrix with
\begin{equation}
    \Delta H_{\pm} \equiv \oint\limits_{\mathcal{C}_{\pm}}
        \rd{H}{t}\;\mathrm{d}t.\label{eq:def_dHpm}
\end{equation}
This can be simplified by using:
\begin{align}
    \rd{H}{t} &=
            \pd{H}{(\cos \theta)}\rd{(\cos \theta)}{t}
            + \pd{H}{\phi}\rd{\phi}{t},\nonumber\\
        &= \p{\rd{(\cos\theta)}{t}}_{\rm tide} \rd{\phi}{t},\\
    \Delta H_{\pm} &= \mp\frac{1}{t_{\rm s}}
        \int\limits_0^{2\pi} \sin^2\theta\;\mathrm{d}\phi.
\end{align}
Thus, if we evaluate $H$ every time that a trajectory crosses $\phi = 0$, we see
that it will initially be negative and increase for each circulation cycle until
the system encounters the separatrix. At the beginning of the
separatrix-crossing orbit, the initial value of $H$, denoted $H_{\rm i}$,
satisfies $H_{\rm sep} - \Delta H_- \leq H_{\rm i} \leq H_{\rm sep}$: if it is
less than the lower bound, it will not cross the separatrix on this orbit, and
if it is greater than the upper bound then it is already inside the separatrix.
The two endpoints of this range are denoted in Fig.~\ref{fig:6equils} by the
black dot and cross at the left and right edges of the plot.

@@

    \item If $H_{\rm f} < H_{\rm sep}$, then the trajectory exits the separatrix
        above CS4, following the yellow shaded region in Fig.~\ref{fig:6equils}
        and executes a zone III $\to$ I transition.
\end{itemize}
Since $H_{\rm f} = H_{\rm i} + \Delta H_+ + \Delta H_-$, we find that if $H_{\rm
i}$ is in the interval $\s{H_{\rm sep} - \Delta H_-, H_{\rm sep} - \Delta H_{-}
- \Delta H_+}$, then the system executes a III $\to$ I transition, and if it
is in the interval $\s{H_{\rm sep} - \Delta H_- - \Delta H_+, H_{\rm sep}}$,
then the system executes a III $\to$ II transition. The values of $\cos
\theta$ for which $H$ is equal to $H_{\rm sep} - \Delta H_-$, $H_{\rm sep} - \Delta
H_- - \Delta H_+$, and $H_{\rm sep}$ (CS4) for $\phi = 0$ are shown in
Fig.~\ref{fig:6equils} as the black cross, star, and dot at $\phi = 0$
respectively. Finally, if the alignment torque is weak, then $\Delta H_- \sim
\mathcal{O}\p{t_{\rm s}^{-1}}$ is small compared to the characteristic values of
$H$, and $H_{\rm i}$ can be effectively considered as uniformly distributed over
$\s{H_{\rm sep} - \Delta H_-, H_{\rm sep}}$. As such, we obtain that
\begin{equation}
    P_{\rm III \to II} = \frac{\Delta H_- + \Delta H_+}{\Delta H_-}.
        \label{eq:def_P32_toy}
\end{equation}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/6manifolds0_20.png}
    \caption{Plot illustrating the probabilistic origin of separatrix capture
    for $\eta = 0.2$, $I = 20^\circ$, and $\abs{gt_{\rm s}} = 10^{-3}$. Orange
    regions converge to CS1, and green to CS2, while CS4 is labeled with the
    purple dots. The boundaries separating the CS1 and CS2-approaching regions
    consist of the critical trajectories (labeled in the legend) that, when
    evolved either forwards or backwards in time (superscripts in the legend
    labels), asymptotically approach CS4 where either $\phi = 0^\circ$ or $\phi
    = 360^\circ$ (subscripts in the legend labels). Within zone III, the regions
    of phase space reaching CS1 and CS2 both become very thin (shown
    qualitatively as the light green lines of decreasing width), reflecting the
    fact that the outcome for a particular initial condition can be approximated
    as probabilistic sufficiently far from the separatrix. Labeled in the blue
    and black dots at the left edge of the plot are the two backwards-in-time
    critical trajectories intersect $\phi = 0$, and are where $H$ is equal to
    $H_{\rm sep} - \Delta H_- - \Delta H_+$ and $H_{\rm sep} - \Delta H_-$
    respectively (Eqs.~\ref{eq:def_Hsep} and~\ref{eq:def_dHpm}).
    }\label{fig:toy_hop_manifolds}
\end{figure}

To actually evaluate Eq.~\eqref{eq:def_P32_toy}, we use the parameterization for
the separatrix \citep{su2020} for $\eta \ll 1$:
\begin{equation}
    \p{\cos \theta}_{\mathcal{C}_{\pm}} \approx
        \eta \cos I \pm \sqrt{2\eta\sin I\p{1 - \cos \phi}}.
        \label{eq:sep_theta}
\end{equation}
It can then be shown that
\begin{align}
    \Delta H_- &\approx \frac{2\pi}{t_{\rm s}}\p{1
        - 2\eta \sin I} + \mathcal{O}(\eta^{3/2}),\\
    \Delta H_+ + \Delta H_- &\approx
        \frac{32 \eta^{3/2}\cos I \sqrt{\sin I}}{t_{\rm s}}
            + \mathcal{O}\p{\eta^{5/2}},\\
    P_{\rm III \to II} &\approx
        \frac{16 \eta^{3/2} \cos I \sqrt{\sin I}}{\pi
            \p{1  - 2\eta \sin I}}.\label{eq:P32_toy}
\end{align}

@@

In Section~\ref{ss:toy_outcomes}, we found that keeping track of the value of
$H$, the value of the
unperturbed Hamiltonian, allowed us to calculate the probabilities of the
various outcomes of separatrix encounter. Specifically, the outcome upon separatrix
encounter is determined by the value of $H$ at the start of the
separatrix-crossing orbit relative to $H_{\rm sep}$, the value of $H$ along the
separatrix. When the spin is also evolving, $H_{\rm sep}$ is also
changing during the separatrix-crossing orbit, and the discussion in
Section~\ref{ss:toy_outcomes} must be generalized to account for this. Instead
of focusing on the evolution of $H$ along a trajectory, we instead follow the
evolution of
\begin{equation}
    K \equiv H - H_{\rm sep}.
\end{equation}
Note that $K > 0$ inside the separatrix, and $K < 0$ outside. Then, the outcome
of the separatrix-crossing orbit is largely the same as discussed in
Section~\ref{ss:toy_outcomes}. First, we must compute the change in $K$ along
the legs of the separatrix. We define $\Delta K_{\pm}$ by generalizing
Eq.~\eqref{eq:def_dHpm} very naturally:
\begin{align}
    \Delta K_{\pm} &= \oint_{\mathcal{C}_{\pm}} \rd{H}{t}
        - \rd{H_{\rm sep}}{t}\;\mathrm{d}t.\label{eq:def_dK_weaktide}
\end{align}
Here, however, note that $\mathcal{C}_{\pm}$ depends on the value of
$\Omega_{\rm s}$ upon separatrix encounter. Since there is no closed form
solution for $\Omega_{\rm s}(t)$, the probabilities of the various outcomes can
only be determined as a function of the system properties at separatrix
encounter, and not as a simple function of the initial conditions.

>>>>

In this section, we consider a simplified model of equilibrium tides that
isolates the important new phenomenon presented in this paper. We assume that
the spin magnitude of the planet is constant, so $\alpha$ and $g$ are both
fixed, while the spin orientation $\uv{s}$ experiences an alignment torque
towards $\uv{l}$ on the alignment timescale $t_{\rm al}$:
\begin{equation}
    \p{\rd{\uv{s}}{t}}_{\rm tide}
        = \frac{1}{t_{\rm al}} \uv{s} \times \p{\uv{l} \times \uv{s}}.
        \label{eq:dsdt_tide_toy}
\end{equation}
The full equations of motion for $\uv{s}$ in the coordinates $\theta$ and $\phi$
can be written:
\begin{align}
    \rd{\theta}{t} &= -g\sin I \sin \phi - \frac{1}{t_{\rm al}} \sin \theta,
        \label{eq:dqdt_toy}\\
    \rd{\phi}{t} &= -\alpha \cos\theta
        - g\p{\cos I + \sin I \cot \theta \cos \phi}.\label{eq:dfdt_toy}
\end{align}

@@

If the alignment torque is weak ($\abs{g}t_{\rm al} \gg 1$), then the fixed
points of Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) are just slightly shifted
CSs. This shift can be calculated cleanly to leading order: all of the CS
obliquities $\theta_{\rm cs}$ are unchanged while the azimuthal angle $\phi_{\rm
cs}$ for each CS now satisfies
\begin{equation}
    \sin \phi_{\rm cs} = \frac{\sin\theta_{\rm cs}}{\sin I \abs{g}t_{\rm al}}.
        \label{eq:mcs_shift}
\end{equation}
We can see that if $t_{\rm al} > t_{\rm al, c}$, where
\begin{align}
    t_{\rm al, c} &= \frac{1}{\abs{g}\sin I},\label{eq:mcs_shift_crit}
\end{align}
then Eq.~\eqref{eq:mcs_shift} will always have solutions for $\phi_{\rm cs}$,
and the alignment torque never changes the number of fixed points of the system.
If $t_{\rm al}$ is decreased below $t_{\rm al, c}$, CS2 and CS4 cease to be
fixed points if $\eta \ll 1$ \citep[as first noted in][]{fabrycky_otides}, as
$\theta_{\rm cs} \approx 90^\circ$ for these (see Fig.~\ref{fig:cs_locs}), while
the other CSs have small $\sin \theta_{\rm cs}$ and are only slightly perturbed.
Figure~\ref{fig:mcs} shows the obliquity and azimuthal angles for each of the
CSs in the $\eta \ll 1$ case, where it can be seen that CS2 and CS4 collide and
annihilate. The phase shift $\phi_{\rm cs}$ for CS2 and CS4 for $t_{\rm al} >
t_{\rm al, c}$ can be predicted to good accuracy using Eq.~\eqref{eq:mcs_shift}
using $\theta_{\rm cs} \approx \pi/2 - \eta \cos (I) \approx 79^\circ$
\citep{su2020} for both, shown as the dashed lines in the bottom panel of
Fig.~\ref{fig:mcs}. For the remainder of this section, we will consider the case
where $t_{\rm al} \gg t_{\rm al, c}$ and the CSs only differ slightly from their
unperturbed locations.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/0_stab.png}
    \caption{Modified CS obliquities (top) and azimuthal angles (bottom) for $I
    = 20^\circ$ and $\eta = 0.2$, where the CS1 and CS3 obliquities have been
    offset (labeled in top legend) to improve clarity of the plot. In both
    panels, the solid lines give the result when applying a numerical root
    finding algorithm to the full equations of motion,
    Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}), while the dotted lines in the
    bottom panel give the CS2 and CS4 azimuthal angles according to
    Eq.~\eqref{eq:mcs_shift}. At $\abs{gt_{\rm al} \sin I} = 1$, CS2 and CS4
    collide and annihilate (Eq.~\ref{eq:mcs_shift_crit}).}\label{fig:mcs}
\end{figure}

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the weak tidal alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a shifted CS, yelding
\begin{align}
    \rd{}{t}\begin{bmatrix}
        \theta\\ \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm al}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and $\Delta \phi = \phi - \phi_{\rm cs}$. The eigenvalues
$\lambda$ of Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{equation}
    0 = \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm al}}}\lambda
        - \lambda_0^2,\label{eq:lambda_orig}
\end{equation}
where
\begin{equation}
    \lambda_0^2 \equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}}\label{eq:def_l0_sq}.
\end{equation}
When $t_{\rm al}$ is large, we can simplify Eq.~\eqref{eq:lambda_orig} to
\begin{equation}
    \lambda \approx -\frac{\cos \theta_{\rm cs}}{t_{\rm al}}
        \pm \sqrt{\lambda_0^2}. \label{eq:lambda_approx}
\end{equation}
The stability depends only on the real part of $\lambda$ in
Eq.~\eqref{eq:lambda_approx}. $\lambda_0^2$ is a generalization of Eq.~(A4)
in, \citet{su2020} and generally has the same behavior: it is negative for CSs
1--3 and positive for CS4, as shown in Fig.~\ref{fig:lambda_full}. Thus, CS4 is
always unstable, as there will always be at least one positive solution for
$\lambda$, and the stability of CSs 1--3 are solely determined by the sign of
$\cos \theta_{\rm cs}$. Evaluating at each of the CSs (see
Fig.~\ref{fig:cs_locs}), we conclude that CS1 and CS2 are stable and attracting
while CS3, stable in the absence of the torque, becomes unstable. These
calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal, fabrycky_otides}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_lambdas_full.png}
    \caption{$\lambda_0^2$ (Eq.~\ref{eq:def_l0_sq}) as a function of $\eta$ for
    the four CSs. The solid lines give $\lambda_0^2$ for the unperturbed
    $\phi_{\rm cs}$, and the dashed lines give the values for $\phi_{\rm cs}$
    shifted by $60^\circ$ ($\phi_{\rm cs} = 120^\circ$ for CS2 and $\phi_{\rm
    cs} = 60^\circ$ for CSs 1, 3, and 4).
    }\label{fig:lambda_full}
\end{figure}

@@

With the above results, we are equipped to ask questions about the dynamics of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}): what is the long term behavior for
a general initial $\uv{s}$? If $\eta > \eta_{\rm c}$, then the only possible
final spin state is CS2, and all initial conditions will evolve asympotically
towards it. As such, we consider only $\eta < \eta_{\rm c}$, where an initial
condition can asymptotically evolve towards either CS1 or CS2. We numerically
integrate Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) for many random initial
conditions uniformly distributed in $\p{\cos \theta, \phi}$ with $\abs{g}t_{\rm
al}= 10^{-3}$ and record the nearest CS for each integration after $10t_{\rm al}$. In Fig.~\ref{fig:toy_phop}, we show the results of this procedure for $\eta
= 0.2$, and $I = 20^\circ$. It is clear that initial conditions in zone I evolve
into CS1, those in zone II evolve into CS2, while those in zone III have a
probabilistic outcome. We aim to understand each of these in turn:
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_2.png}
    \caption{Plot illustrating the asymptotic behavior of initial conditions for
    $\eta = 0.2$ and $I = 20^\circ$. Each dot represents an initial spin
    orientation, and the coloring of the dot indicates which Cassini State
    (legend) the system asymptotes towards.}\label{fig:toy_phop}
\end{figure}

@@

To understand this process quantitatively, as well as associate probabilities to
the two possible outcomes, it is important to be more quantitative. The correct
approach is to consider the evolution of the value of the \emph{unperturbed}
Hamiltonian (Eq.~\ref{eq:H}) as the spin evolves with the alignment torque. A
point in zone III evolves such that $H$ is increasing until $H \approx H_{\rm
sep}$ where $H_{\rm sep}$ is the value of $H$ along the separatrix, given by
\begin{align}
    H_{\rm sep} &\equiv H\p{\cos \theta_{\rm 4}, \phi_{\rm 4}},\nonumber\\
        &\approx g\sin I + \frac{g^2}{2\alpha}\cos^2 I +
            \mathcal{O}\p{\eta^2},\label{eq:def_Hsep}
\end{align}
where $\theta_4$ and $\phi_4$ are the coordinates of CS4. As the system evolves
closer to the separatrix, the change in $H$ over each circulation cycle can be
approximated by $\Delta H_-$, the change in $H$ along $\mathcal{C}_-$ (see
Fig.~\ref{fig:1contours}). In general, we can compute $\Delta H_{\pm}$ the
change over both legs of the separatrix with
\begin{equation}
    \Delta H_{\pm} \equiv \oint\limits_{\mathcal{C}_{\pm}}
        \rd{H}{t}\;\mathrm{d}t.\label{eq:def_dHpm}
\end{equation}
This can be simplified by using:
\begin{align}
    \rd{H}{t} &=
            \pd{H}{(\cos \theta)}\rd{(\cos \theta)}{t}
            + \pd{H}{\phi}\rd{\phi}{t},\nonumber\\
        &= \p{\rd{(\cos\theta)}{t}}_{\rm tide} \rd{\phi}{t},\\
    \Delta H_{\pm} &= \mp\frac{1}{t_{\rm al}}
        \int\limits_0^{2\pi} \sin^2\theta\;\mathrm{d}\phi.
\end{align}
Thus, if we evaluate $H$ every time that a trajectory crosses $\phi = 0$, we see
that it will initially be negative and increase for each circulation cycle until
the system encounters the separatrix. At the beginning of the
separatrix-crossing orbit, the initial value of $H$, denoted $H_{\rm i}$,
satisfies $H_{\rm sep} - \Delta H_- \leq H_{\rm i} \leq H_{\rm sep}$: if it is
less than the lower bound, it will not cross the separatrix on this orbit, and
if it is greater than the upper bound then it is already inside the separatrix.
The two endpoints of this range are denoted in Fig.~\ref{fig:6equils} by the
black dot and cross at the left and right edges of the plot.

@@

    \item If $H_{\rm f} < H_{\rm sep}$, then the trajectory exits the separatrix
        above CS4, following the yellow shaded region in Fig.~\ref{fig:6equils}
        and executes a zone III $\to$ I transition.
\end{itemize}
Since $H_{\rm f} = H_{\rm i} + \Delta H_+ + \Delta H_-$, we find that if $H_{\rm
i}$ is in the interval $\s{H_{\rm sep} - \Delta H_-, H_{\rm sep} - \Delta H_{-}
- \Delta H_+}$, then the system executes a III $\to$ I transition, and if it
is in the interval $\s{H_{\rm sep} - \Delta H_- - \Delta H_+, H_{\rm sep}}$,
then the system executes a III $\to$ II transition. The values of $\cos
\theta$ for which $H$ is equal to $H_{\rm sep} - \Delta H_-$, $H_{\rm sep} - \Delta
H_- - \Delta H_+$, and $H_{\rm sep}$ (CS4) for $\phi = 0$ are shown in
Fig.~\ref{fig:6equils} as the black cross, star, and dot at $\phi = 0$
respectively. Finally, if the alignment torque is weak, then $\Delta H_- \sim
\mathcal{O}\p{t_{\rm al}^{-1}}$ is small compared to the characteristic values of
$H$, and $H_{\rm i}$ can be effectively considered as uniformly distributed over
$\s{H_{\rm sep} - \Delta H_-, H_{\rm sep}}$. As such, we obtain that
\begin{equation}
    P_{\rm III \to II} = \frac{\Delta H_- + \Delta H_+}{\Delta H_-}.
        \label{eq:def_P32_toy}
\end{equation}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/6manifolds0_20.png}
    \caption{Plot illustrating the probabilistic origin of separatrix capture
    for $\eta = 0.2$, $I = 20^\circ$, and $\abs{gt_{\rm al}} = 10^{-3}$. Orange
    regions converge to CS1, and green to CS2, while CS4 is labeled with the
    purple dots. The boundaries separating the CS1 and CS2-approaching regions
    consist of the critical trajectories (labeled in the legend) that, when
    evolved either forwards or backwards in time (superscripts in the legend
    labels), asymptotically approach CS4 where either $\phi = 0^\circ$ or $\phi
    = 360^\circ$ (subscripts in the legend labels). Within zone III, the regions
    of phase space reaching CS1 and CS2 both become very thin (shown
    qualitatively as the light green lines of decreasing width), reflecting the
    fact that the outcome for a particular initial condition can be approximated
    as probabilistic sufficiently far from the separatrix. Labeled in the blue
    and black dots at the left edge of the plot are the two backwards-in-time
    critical trajectories intersect $\phi = 0$, and are where $H$ is equal to
    $H_{\rm sep} - \Delta H_- - \Delta H_+$ and $H_{\rm sep} - \Delta H_-$
    respectively (Eqs.~\ref{eq:def_Hsep} and~\ref{eq:def_dHpm}).
    }\label{fig:toy_hop_manifolds}
\end{figure}

To actually evaluate Eq.~\eqref{eq:def_P32_toy}, we use the parameterization for
the separatrix \citep{su2020} for $\eta \ll 1$:
\begin{equation}
    \p{\cos \theta}_{\mathcal{C}_{\pm}} \approx
        \eta \cos I \pm \sqrt{2\eta\sin I\p{1 - \cos \phi}}.
        \label{eq:sep_theta}
\end{equation}
It can then be shown that
\begin{align}
    \Delta H_- &\approx \frac{2\pi}{t_{\rm al}}\p{1
        - 2\eta \sin I} + \mathcal{O}(\eta^{3/2}),\\
    \Delta H_+ + \Delta H_- &\approx
        \frac{32 \eta^{3/2}\cos I \sqrt{\sin I}}{t_{\rm al}}
            + \mathcal{O}\p{\eta^{5/2}},\\
    P_{\rm III \to II} &\approx
        \frac{16 \eta^{3/2} \cos I \sqrt{\sin I}}{\pi
            \p{1  - 2\eta \sin I}}.\label{eq:P32_toy}
\end{align}

@@

In Section~\ref{ss:toy_outcomes}, we found that keeping track of the value of
$H$, the value of the
unperturbed Hamiltonian, allowed us to calculate the probabilities of the
various outcomes of separatrix encounter. Specifically, the outcome upon separatrix
encounter is determined by the value of $H$ at the start of the
separatrix-crossing orbit relative to $H_{\rm sep}$, the value of $H$ along the
separatrix. When the spin is also evolving, $H_{\rm sep}$ is also
changing during the separatrix-crossing orbit, and the discussion in
Section~\ref{ss:toy_outcomes} must be generalized to account for this. Instead
of focusing on the evolution of $H$ along a trajectory, we instead follow the
evolution of
\begin{equation}
    K \equiv H - H_{\rm sep}.
\end{equation}
Note that $K > 0$ inside the separatrix, and $K < 0$ outside. Then, the outcome
of the separatrix-crossing orbit is largely the same as discussed in
Section~\ref{ss:toy_outcomes}. First, we must compute the change in $K$ along
the legs of the separatrix. We define $\Delta K_{\pm}$ by generalizing
Eq.~\eqref{eq:def_dHpm} in the natural way:
\begin{align}
    \Delta K_{\pm} &= \oint_{\mathcal{C}_{\pm}} \rd{H}{t}
        - \rd{H_{\rm sep}}{t}\;\mathrm{d}t.\label{eq:def_dK_weaktide}
\end{align}
Here, however, note that $\mathcal{C}_{\pm}$ depends on the value of
$\Omega_{\rm s}$ upon separatrix encounter. Since there is no closed form
solution for $\Omega_{\rm s}(t)$, the probabilities of the various outcomes
cannot be expressed as a simple function of the initial conditions.

Then, if we call $K_i$ the value of $K$ at the start ($\phi = 0$) of the
separatrix-crossing orbit, $K_i > -\Delta K_+ - \Delta K_-$ gives a III
$\to$ II transition and eventual evolution towards tCE2 while $-\Delta K_- < K_i
< -\Delta K_- - \Delta K_+$ gives a III $\to$ I transition and ultimate
evolution towards tCE1. Thus, we find that the probability of a III $\to$ II
transition is given by
\begin{equation}
    P_{\rm III \to II} = \frac{\Delta K_+ + \Delta K_-}{\Delta
        K_-}.\label{eq:def_pc_weaktide}
\end{equation}
Since $\Delta K_{\pm}$ are evaluated at resonance encounter, and
$\Omega_{\rm s}$ is evolving, there is no way to express $\Delta K_{\pm}$ as a
closed form of initial conditions. In fact, since many resonance encounters
occur when $\eta$ is substantial, even an approximate calculation using
Eq.~\eqref{eq:sep_theta} is inaccurate, and we instead calculate $\Delta
K_{\pm}$ along the two legs of the separatrix obtained numerically via a root
finding algorithm. Note that
Eqs.~(\ref{eq:def_dK_weaktide},~\ref{eq:def_pc_weaktide}) are equivalent to the
separatrix capture result of \citet{henrard1982} when $\dot{\theta}_{\rm tide} =
0$ \citep{henrard1987}. Here, we have argued that this classic calculation can
be unified with the calculation given in Section~\ref{ss:toy_outcomes} to give
an accurate prediction of separatrix encounter outcome probabilities in the
presence of both a dissipative perturbation and a parametric variation of the
Hamiltonian.
